---
title: "Dataset: IBD cosMX GarridoTrigo2023"
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---


Data is from paper [_Macrophage and neutrophil heterogeneity at single-cell spatial resolution in human inflammatory bowel disease_](https://www.ncbi.nlm.nih.gov/pubmed/37495570)
from Garrido-Trigo et al 2023, [@garrido-trigoMacrophageNeutrophilHeterogeneity2023]

The study included 9 cosmx slides of colonic biopsies

* 3x HC - Healthy controls
* 3x UC - Ulcerative colitis
* 3x CD - Chrones's disease


Fastantically - not only have they made their raw and annotated data available, but have also shared their analysis code; https://github.com/HelenaLC/CosMx-SMI-IBD


They have also shared browseable interface here: https://servidor2-ciberehd.upc.es/external/garrido/app/


# Libraries

```{r}
library(SingleCellExperiment)
library(tidyverse)
```

# Data download

From the project description:

Raw data on GEO here; https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE234713

_Inflammatory bowel diseases (IBDs) including ulcerative colitis (UC) and Crohn’s disease (CD) are chronic inflammatory diseases with increasing worldwide prevalence that show a perplexing heterogeneity in manifestations and response to treatment. We applied spatial transcriptomics at single-cell resolution (CosMx Spatial Molecular Imaging) to human inflamed and uninflamed intestine._


The following files were download from GEO

* GSE234713_CosMx_annotation.csv.gz
* GSE234713_CosMx_normalized_matrix.txt.gz
* GSE234713_RAW.tar: RAW data was downloaded via custom downlod in 3 batches, one per group
  * GSE234713_RAW_CD.tar
  * GSE234713_RAW_HC.tar
  * GSE234713_RAW_UC.tar
* GSE234713_ReadMe_SMI_Data_File.html

```{sh eval=FALSE}
cd raw_data
tar -xf GSE234713_RAW_CD.tar
tar -xf GSE234713_RAW_HC.tar
tar -xf GSE234713_RAW_UC.tar
```



# Data load


```{r}
dataset_dir      <- '~/projects/spatialsnippets/datasets/'
project_data_dir <- file.path(dataset_dir,'GSE234713_IBDcosmx_GarridoTrigo2023')
sample_dir            <- file.path(project_data_dir, "raw_data/") 
annotation_file       <- file.path(project_data_dir,"GSE234713_CosMx_annotation.csv.gz")
data_dir              <- file.path(project_data_dir, "processed_data/") 


seurat_file_01_loaded <- file.path(data_dir, "GSE234713_CosMx_IBD_seurat_01_loaded.RDS")

# config
min_count_per_cell <- 100
max_pc_negs        <- 1.5
max_avg_neg        <- 0.5



sample_codes <- c(HC="Healthy controls",UC="Ulcerative colitis",CD="Crohn's disease")

 

```



```{r}
################################################################################
# SFE



load_and_save_initial_sfe_to_disk <- function(sfe_data_dirs, the_samples, sfe_out_dir) {
  
  if (dir.exists(sfe_out_dir)) {
    stop("SFE out directory already exists, please remove")
  }
  stopifnot(length(sfe_data_dirs) == length(the_samples))
  
  # Setup first or only slide
  sfe <- load_sfe_with_molecules(sfe_data_dirs[1], the_samples[1]) 
  sfe <- add_basic_qc_annontation(sfe)
  
  
  
  # Handle multiple samples
  if (length(sfe_data_dirs) > 1) {
    for (n in 2:length(sfe_data_dirs) ) {
      sfe2 <- load_sfe_with_molecules(sfe_data_dirs[n], the_samples[n]) 
      sfe2 <- add_basic_qc_annontation(sfe2) # Qc per slide.
      # There's gotta be a more effient way than cbind? But we only have 2.
      sfe <- cbind(sfe, sfe2)
    }
  }
  
  # Add probe stats After merge, since there's one stat per probe.
  sfe <- add_basic_probe_stats(sfe)
  
  # Save using the method from alabaster.spatioal.
  #library(alabaster.bumpy)
  saveObject(sfe, sfe_out_dir )
  return(sfe_out_dir)
  
}

load_sfe_with_molecules <- function (sfe_data_dir, the_sample, sfe_out_dir) {
  

  # Where add_molecules=TRUE.
  #Error: Capacity error: array cannot contain more than 2147483646 bytes, have 2148177769
  # Pulls cell ids from cell_id
  #"1_1" "2_1" "3_1" "4_1" "5_1" "6_1"
  # "165_349" "166_349" "167_349" "168_349" "169_349" "170_349"
  # <cellnum>_<fov>
  sfe  <- readCosMX.edit(data_dir=sfe_data_dir, 
                    sample_id = the_sample,
                      add_molecules = FALSE # True yeilds errors.
                      )
  ###sfe$sample_id  <- the_sample #its not really a sample_id,  but this is added in load
  sfe$slide_name <- the_sample # its a nice slide name.
  
  
  # the tx file within the data dir
  tx_file <- file.path(sfe_data_dir, paste0(the_sample,"_tx_file.csv"))
  # Read in transcript coordinates, but only keep cellular tx (strips out alot)
  #fov cell_ID      cell x_local_px y_local_px x_global_px y_global_px     z target  CellComp
  #1       1   c_1_1_1       4255        140   10366.964    126795.1     2   Cd74 Cytoplasm
  #1       1   c_1_1_1       4255        140   10366.944    126795.5     7 Ifitm3 Cytoplasm
  #1       1   c_1_1_1       4252        139   10363.784    126796.5     6  Acta2 Cytoplasm
  #1       1   c_1_1_1       4250        150   10362.394    126785.8     3  Ifit1 Cytoplasm
  mol_table <- fread(tx_file, 
                     #nrow=100000,
                     select=c('cell_ID', 'fov', 'target', 'x_global_px', 'y_global_px', 'CellComp')) 
  dim(mol_table)
  mol_table <- mol_table[CellComp != 'None', ] # Assigned to cells only
  mol_table <- mol_table[,cell_id:=paste0(cell_ID,"_",fov)]
  mol_table <- mol_table[,!c('CellComp', "fov","cell_ID")]
  dim(mol_table)
  
  
  # construct BumpyMatrix
  mol_bm <- splitAsBumpyMatrix(
    mol_table[, c("x_global_px", "y_global_px")], 
    row = mol_table$target, col = mol_table$cell_id,
    sparse =TRUE) # sparse=TRUE is important, and not default !! 10fold less ram.
  rm(mol_table)
  
  
  # there may be more 'cells' from the tx file than in the object, but these 
  # are quietly ignored. But everything in sft should be in mol_bm!
  stopifnot(all(rownames(sfe) %in% rownames(mol_bm) ))
  stopifnot(all( colnames(sfe) %in% colnames(mol_bm) ))
  #mol_bm <- mol_bm[,colnames(sfe)] # match order and drop cells not in sfe
  #mol_bm <- mol_bm[rownames(sfe),] # ditto genes
  mol_bm <- mol_bm[rownames(sfe),colnames(sfe)] # match order and drop cells not in sfe, ditto genes
  
 
  # now save the molecules in the object
  assay(sfe, 'molecules') <- mol_bm
  # molecules(sfe)['Sell',1])

  # Where there is missing polygon data for sfe, 
  # remove the corresponding cell
  # Otherwise you end up with spatial annotations instead of segmentations,
  # which means files cant be joined (and probably other issues!)
  # Affects one (1!) cell in MulscleX110, I do not care to see why.
  # Should do nothing otherwise.
  sfe<-check_and_rm_cells_without_polygons(sfe)
  
  return(sfe)  
}



```
