---
title: "Dataset: IBD cosMX GarridoTrigo2023"
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---


Data is from paper [_Macrophage and neutrophil heterogeneity at single-cell spatial resolution in human inflammatory bowel disease_](https://www.ncbi.nlm.nih.gov/pubmed/37495570)
from Garrido-Trigo et al 2023, [@garrido-trigoMacrophageNeutrophilHeterogeneity2023]

The study included 9 cosmx slides of colonic biopsies

* 3x HC - Healthy controls
* 3x UC - Ulcerative colitis
* 3x CD - Chrones's disease


Fastantically - not only have they made their raw and annotated data available, but have also shared their analysis code; https://github.com/HelenaLC/CosMx-SMI-IBD


They have also shared browseable interface here: https://servidor2-ciberehd.upc.es/external/garrido/app/


# Libraries

```{r}

library(tidyverse)
library(patchwork)
library(SpatialExperiment)
library(SpatialExperimentIO) #reading functions
library(alabaster.spatial) # SaveObject
library(scran)  
library(scater)
library(scuttle)
library(bluster) # clustering parameters, needed for passing htreads to NNGraphParam
library(BiocParallel)


#library(SpatialFeatureExperiment)
#library(alabaster.sfe)
# Requires v1.9.3 to use alabaster.sfe to save object.
# alabaster.sfe not yet in bioconductor production March 2025
#renv::install('pachterlab/SpatialFeatureExperiment@devel')
#renv::install('pachterlab/alabaster.sfe')

```

# Config

```{r}
dataset_dir      <- '~/projects/spatialsnippets/datasets/'
project_data_dir <- file.path(dataset_dir,'GSE234713_IBDcosmx_GarridoTrigo2023')
sample_dir            <- file.path(project_data_dir, "raw_data_for_sfe/") 
annotation_file       <- file.path(project_data_dir,"GSE234713_CosMx_annotation.csv.gz")
data_dir              <- file.path(project_data_dir, "processed_data/") 


sfe_01_loaded <- file.path(data_dir, "GSE234713_CosMx_IBD_sfe_01_loaded")

# config
min_count_per_cell <- 50 # previusly 100
min_detected_genes_per_cell <- 20 # from paper.
max_pc_negs        <- 1.5
max_avg_neg        <- 0.5

sample_codes <- c(HC="Healthy controls",UC="Ulcerative colitis",CD="Crohn's disease")

```

# Data download

From the project description:

Raw data on GEO here; https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE234713

_Inflammatory bowel diseases (IBDs) including ulcerative colitis (UC) and Crohn’s disease (CD) are chronic inflammatory diseases with increasing worldwide prevalence that show a perplexing heterogeneity in manifestations and response to treatment. We applied spatial transcriptomics at single-cell resolution (CosMx Spatial Molecular Imaging) to human inflamed and uninflamed intestine._


The following files were download from GEO

* GSE234713_CosMx_annotation.csv.gz
* GSE234713_CosMx_normalized_matrix.txt.gz
* GSE234713_RAW.tar: RAW data was downloaded via custom downlod in 3 batches, one per group
  * GSE234713_RAW_CD.tar
  * GSE234713_RAW_HC.tar
  * GSE234713_RAW_UC.tar
* GSE234713_ReadMe_SMI_Data_File.html

```{sh eval=FALSE}
cd raw_data
tar -xf GSE234713_RAW_CD.tar
tar -xf GSE234713_RAW_HC.tar
tar -xf GSE234713_RAW_UC.tar
```

Construct a folder with each slide's info.

NB: Also due to limited temp directory space, I need to gunzip the files (this really shouldn't be needed)
that get read by the efficient DT reading functions. Symptom of that is silently incomplete file reads.
https://www.linkedin.com/pulse/trivial-fix-after-3-hours-debugging-kirill-tsyganov/


```{sh eval=FALSE}
mkdir raw_data_for_sfe
cp raw_data/*tar.gz raw_data_for_sfe/
cd raw_data_for_sfe/
tar -xzf *.tar.gz


cd /home/s.williams/projects/spatialsnippets/datasets/GSE234713_IBDcosmx_GarridoTrigo2023/raw_data_for_sfe
for sample in GSM7473682_HC_a  GSM7473685_UC_a  GSM7473688_CD_a GSM7473683_HC_b  GSM7473686_UC_b  GSM7473689_CD_b GSM7473684_HC_c  GSM7473687_UC_c  GSM7473690_CD_c; do
  echo $sample
  #cd /home/s.williams/projects/spatialsnippets/datasets/GSE234713_IBDcosmx_GarridoTrigo2023/raw_data_for_sfe
  
  mkdir raw_data_for_sfe/${sample}
  cp raw_data/${sample}_* raw_data_for_sfe/${sample}
  gunzip raw_data_for_sfe/${sample}/*
  
  # Separately copy in the polygons files (supplied by authors)
  samplecode=${sample: -4}
  echo $samplecode
  cp /home/s.williams/projects/spatialsnippets/datasets/GSE234713_IBDcosmx_GarridoTrigo2023/polygons/${samplecode}.csv ${sample}/${sample}-polygons.csv
  
  
  
done 


```


Read in each sample as a 'SpatialExperiment' using a function from the *SpatialExperimentIO* package.

We don't have polygon files for this dataset, otherwise would load a a SpatialFeatureExperiment. A spatialFeatureExperiment object *is a* SpatialExperiment object with extra features (ie. it inherits from it).






# Data load


Function to load and annotate one sample
```{r}
#the_sample   = 'GSM7473682_HC_a'
#sfe_data_dir = file.path(sample_dir, the_sample )

load_one_sample_as_spe <- function(sfe_data_dir, the_sample) {
  # Note there is no polygon file.
  # Can't real with readCosmX function in this case. (there's no 'feature')
  #GSM7473682_HC_a_exprMat_file.csv  
  #GSM7473682_HC_a_fov_positions_file.csv  
  #GSM7473682_HC_a_metadata_file.csv
  #GSM7473682_HC_a_tx_file.csv
    
  spe <- readCosmxSXE(sfe_data_dir, addPolygon=FALSE, addTx = TRUE)
  
  # sample code nice and short, and may be inferred
  sample_code <- substr(the_sample,12,16)
  spe$sample_id  <- the_sample
  spe$slide_name <- the_sample

  # and some experimental info
  spe$individual_code <- substr(the_sample,12,16)
  spe$tissue_sample   <- substr(the_sample,12,16)
  spe$group           <- factor(substr(the_sample, 12,13), levels=names(sample_codes))
  spe$condition       <- factor(as.character(sample_codes[spe$group]), levels=sample_codes)
  spe$fov_name        <- paste0(spe$individual_code,"_", str_pad(spe$fov, 3, 'left',pad='0'))
  
  # cell labels: need to match that of the annotation file:
  #                id   subset             SingleR2
  # HC_c_2_1 HC_c_2_1   stroma            Pericytes
  # HC_c_3_1 HC_c_3_1   stroma          Endothelium
  # HC_c_4_1 HC_c_4_1   stroma            Pericytes
  
  
  #Have this:

  #      fov   cell_ID      Area AspectRatio CenterX_local_px
  #1         1         1      6153        0.67             2119
  #2         1        10      5971        0.63             1265
  #3         1       100      1785        1.11             2954
  #4         1      1000      3946        0.59             2959
  
  #     individual_code tissue_sample    group        condition    fov_name
  #                HC_a          HC_a       HC Healthy controls    HC_a_001
  #                HC_a          HC_a       HC Healthy controls    HC_a_001
  #                HC_a          HC_a       HC Healthy controls    HC_a_001
  
  #> table(table(spe$cell_ID))
  #   1    2    3    4    5    6    7    8    9   10   11   12   13   14   15   16   17   18   19 
  # 699  110   56   78   79  110  100  165   21    4   24   38   24    5   66  265   36  190 1451 
  
  
  # HC_c_4_1
  # <>>sample code>_<cell_ID>_<fov
  spe$cell <- paste0(spe$tissue_sample, "_",spe$cell_ID, "_",spe$fov)  

  spe$total_count <- colSums(counts(spe))
  spe$distinct_genes <- colSums(counts(spe)!=0)
  spe$total_count_log10 <- log10(spe$total_count)
  

  #Negative probes  
  spe$neg_count   <- colSums(counts(altExp(spe,'NegPrb')))
  spe$avg_neg     <- colMeans(counts(altExp(spe,'NegPrb')))
  spe$pc_neg<- spe$neg_count / (spe$neg_count + spe$total_count) * 100

  
  return(spe)
}


```


Load an annotate all samples in the directory full of samples (each as flatfiles)
```{r}
# Get list of samples
sample_names <- list.files(sample_dir)
sample_dirs  <- file.path(sample_dir, sample_names)

# Use multiple apply to run each with corresponding path.
spe_list <- mapply(FUN=load_one_sample_as_spe , 
                   sfe_data_dir=sample_dirs, 
                   the_sample= sample_names) 

# This is ineffient, but it works - cbind can join pairs of spe object.
spe <- do.call(cbind, spe_list)

#################################
######## SUBSET 
#spe.full <- spe
#spe <- spe.full[, sample(ncol(spe), size=2000)]
##############################
```


Add the cell annotation from the paper.

```{r}
anno_table <- read_csv(annotation_file)
anno_table <- as.data.frame(anno_table)
rownames(anno_table) <- anno_table$id

head(colData(spe))
head(anno_table)


spe$celltype_subset   <- factor(anno_table[spe$cell,]$subset)
spe$celltype_SingleR2 <- factor(anno_table[spe$cell,]$SingleR2)

# and foactorise a few things now we've got the full table
spe$fov_name          <- factor(spe$fov_name)
spe$individual_code   <- factor(spe$individual_code)
spe$tissue_sample     <- factor(spe$tissue_sample)



# There are a few unannotted, will remove
table(is.na(spe$celltype_subset))
```



# Basic QC filter

## Min count per cell

For interest, not filtering on total counts.
```{r}
ggplot(colData(spe), aes(x=total_count, col=sample_id)) +
  geom_density() + 
  scale_x_log10() +
  theme_bw() +
  ggtitle("Counts per cell")
```

## Min detected genes per cell

Distinct genes observed per cell (number of detected genes per cell)

```{r}
ggplot(colData(spe), aes(x=distinct_genes, col=sample_id)) +
  geom_density() + 
  geom_vline(xintercept = min_detected_genes_per_cell, lty=3) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Counts per cell")
```


## Percent Negative probes

```{r}
ggplot(colData(spe), aes(x=pc_neg, col=sample_id)) +
  geom_density() + 
  geom_vline(xintercept = max_pc_negs, lty=3) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Negative probe composition")
```

```{r}
ggplot(colData(spe), aes(x=avg_neg, col=sample_id)) +
  geom_density() + 
  geom_vline(xintercept = max_avg_neg, lty=3) +
  theme_bw() +
  ggtitle("Negative probe average")
```


Use bottom right corner;
```{r}
ggplot(colData(spe), aes(y=avg_neg, x=total_count)) +
  geom_point(pch=3, alpha=0.1) + 
  geom_hline(yintercept = max_avg_neg, lty=3) +
  geom_vline(xintercept = min_count_per_cell, lty=3) +
  scale_x_log10() + 
  theme_bw() +
  ggtitle("Negative probes vs counts")
```


## Apply filteres

From paper: _"Cells with an average negative control count greater than 0.5 and less than 20 detected features were filtered out."_

Applying a different threshold (arbitrarily), results will differ.


Also remove cell with no cell type annotations.
```{r}
ncol(spe)
spe <- spe[ ,spe$distinct_genes >= min_detected_genes_per_cell &
               spe$avg_neg <= max_avg_neg &
               !(is.na(spe$celltype_subset) )]
ncol(spe)

```



## How many cells per sample?

```{r}
table(spe$tissue_sample)
```



# Basic preprocessing

```{r eval=FALSE}
#hvg_prop=0.3
#num_pcs=20
#sampleblock = 'tissue_sample'
#BPPARAM = MulticoreParam(16)
#spe.full <- spe
#spe <- spe[,sample(ncol(spe),size=5000)]

do_basic_preprocessing <- function(spe, 
                                   hvg_prop = 0.3, 
                                   num_pcs  = 20,
                                   sampleblock = NULL,
                                   num_threads = 16, # used by cluster cells
                                   BPPARAM  = MulticoreParam(num_threads) # and everything else others
                                   ) {
  

  
  # Normalization.
  print('Normalise...')
  spe <- logNormCounts(spe, BPPARAM=BPPARAM)
  
  # Highly variable genes, but ignoring individual level data.
  # Attempts to model technival vs bio variation
  # Choosing 30% of genes, since we've only got 1000, and they are selected for cell type discernment.
  # 30% is arbitrary!
  print("Model gene variance...")
  
  if (is.null(sampleblock)) {
    gene_variances <- modelGeneVar(spe, BPPARAM=BPPARAM) 
  } else {
    to_block <- as.factor(colData(spe)[,sampleblock])
    gene_variances <- modelGeneVar(spe, BPPARAM=BPPARAM, block=to_block) 
  }
  
  print("get top hvg...")
  hvg <- getTopHVGs(gene_variances, prop=hvg_prop)
  print("run pca...")
  spe <- fixedPCA(spe, subset.row = hvg)
  print("run UMAP...")
  spe <- runUMAP(spe, pca=num_pcs, BPPARAM=BPPARAM)
  
  print("cluster...")
  set.seed(12) # set seed for consistant clustering.
  spe$nn.cluster <- clusterCells(spe, use.dimred="PCA", BLUSPARAM = bluster::NNGraphParam(num.threads=16))
  
  
  
  
  spe$cluster_code <- factor(paste0("c",spe$nn.cluster), levels=paste0("c",levels(spe$nn.cluster)))
  
  print("Done")
  return(spe)
  
}

spe <- do_basic_preprocessing(spe, num_pcs = 15)
saveObject(spe, sfe_01_loaded)
```




Load one we prepared earlier.
```{r}
spe <- readObject(sfe_01_loaded)
```



# Basic plots



## UMAP {.tabset .tabset-pills}

### Sample

```{r}
plotUMAP(spe, colour_by = 'tissue_sample', point_shape='.') +
  guides(colour = guide_legend(override.aes = list(shape=15)))
```


### Condition

```{r}
plotUMAP(spe, colour_by = 'condition', point_shape='.') +
  guides(colour = guide_legend(override.aes = list(shape=15)))
```



### Condition X Sample

```{r fig.width=12, fig.height=3}
p1 <- plotUMAP(spe[,spe$group=='HC'], colour_by = 'tissue_sample', point_shape='.') + ggtitle ("Healthy Controls") + 
  guides(colour = guide_legend(override.aes = list(shape=15)))
p2 <- plotUMAP(spe[,spe$group=='UC'], colour_by = 'tissue_sample', point_shape='.') + ggtitle("Ulcerative colitis") +
  guides(colour = guide_legend(override.aes = list(shape=15)))
p3 <- plotUMAP(spe[,spe$group=='CD'], colour_by = 'tissue_sample', point_shape='.') + ggtitle("Crohn's disease")+
  guides(colour = guide_legend(override.aes = list(shape=15)))


p1 + p2 + p3
```

### Total counts

```{r}
plotUMAP(spe, colour_by = 'total_count_log10', point_shape='.') 
```


### Clustered

```{r}
plotUMAP(spe, colour_by = 'cluster_code', point_shape='.', text_by='cluster_code') +
  guides(colour = guide_legend(override.aes = list(shape=15)))
```


### celltype_subset

Classifications from Garrido-Trigo et al 2023.

```{r}
plotUMAP(spe, colour_by = 'celltype_subset', text_by='celltype_subset', point_shape='.') +
  guides(colour = guide_legend(override.aes = list(shape=15)))

```

### celltype_SingleR2

Classifications from Garrido-Trigo et al 2023.

```{r fig.width=12}
plotUMAP(spe, colour_by = 'celltype_SingleR2', point_shape='.') +
  guides(colour = guide_legend(override.aes = list(shape=15)))

```



## Clustering

Double checking the clusters line up with annotaiaon
```{r}
table(spe$celltype_subset)
table(spe$celltype_SingleR2)
table(spe$cluster_code) 
```


Composition. Expect an obvious but imperfect grouping between clustering and cell typing. Will be using annotation for analysis.

```{r}
heatmap(table(spe$celltype_SingleR2, spe$celltype_subset))
heatmap(table(spe$celltype_subset, spe$cluster_code))
heatmap(table(spe$celltype_SingleR2,spe$cluster_code))
```



