---
title: "Dataset: Cosmx Lung "
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

The classic NSLC lung cancer dataset.
https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/nsclc-ffpe-dataset/

Prototype machine. Liver is probably better.

# Libraries

```{r}
library(Seurat)
library(tidyverse)
```

# Data download

Seurat object (with transcript positions) downloaded directly from nanostring website.

https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/human-liver-rna-ffpe-dataset/

NOTE: Download links only work in chrome (welcome to the early 2000s), but you can copy the link from there to wget/curl.

<!-- curl https://smi-public.objects.liquidweb.services/LiverDataReleaseSeurat_newUMAP.RDS  --output LiverDataReleaseSeurat_newUMAP.RDS --> 


Do not use this dataset. Its only really 1 sample so less relevant. 


# Data load

```{r}
dataset_dir <- '/Users/s2992547/data_local/datasets/'
liver_cosmx_seurat_file <- file.path(dataset_dir, 'cosmx_smi_human_liver_ffpe_dataset/LiverDataReleaseSeurat_newUMAP.RDS')
```



```{r eval=FALSE}
so <- readRDS(liver_cosmx_seurat_file )
# Error: vector memory exhausted (limit reached?)
# hits 14G ram. 
```


