---
title: "Dataset: Digits in a dish. "
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Preparation of data from: 
*Digits in a dish: An in vitro system to assess the molecular genetics of hand/foot development at single-cell resolution* Allison M. Fuiten, Yuki Yoshimoto, Chisa Shukunami, H. Scott Stadler. Fronteirs in Cell and Developmental Biology 2023.

https://www.frontiersin.org/articles/10.3389/fcell.2023.1135025/full

Data from GEO, GSE221883. https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE221883


Data processing here is simplified for demonstrative purposes - and differs from that used in the paper!

<!-- Package install
Seurat5
#renv::install('satijalab/seurat-wrappers@seurat5') # Yes.
#renv::install('satijalab/seurat@seurat5')
-->

# Libraries

```{r}
library(Seurat)
library(tidyverse)
```


# Data Download


Download counts matricies from GEO. Note the read10X function used later expects a folder per sample with 
files exactly named barcodes.tsv.gz/features.tsv.gz and matrix.mtx.gz

```{sh eval=FALSE}
wget https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE221883&format=file
tar -xzf GSE221883_RAW.tar

mkdir data_for_seurat



mkdir data_for_seurat
mkdir seurat_objects
for sample in GSM6908653_Day2_A GSM6908655_Day7_A GSM6908657_Day10_A GSM6908654_Day2_B GSM6908656_Day7_B GSM6908658_Day10_B
do
echo ${sample}
mkdir data_for_seurat/${sample}
cp ${sample}_barcodes.tsv.gz data_for_seurat/${sample}/barcodes.tsv.gz
cp ${sample}_features.tsv.gz data_for_seurat/${sample}/features.tsv.gz
cp ${sample}_matrix.mtx.gz data_for_seurat/${sample}/matrix.mtx.gz
done
```




Contains the following files:
```
GSM6908653_Day2_A_barcodes.tsv.gz		
GSM6908653_Day2_A_features.tsv.gz		
GSM6908653_Day2_A_matrix.mtx.gz		
GSM6908654_Day2_B_barcodes.tsv.gz		
GSM6908654_Day2_B_features.tsv.gz		
GSM6908654_Day2_B_matrix.mtx.gz		
GSM6908655_Day7_A_barcodes.tsv.gz		
GSM6908655_Day7_A_features.tsv.gz		
GSM6908655_Day7_A_matrix.mtx.gz		
GSM6908656_Day7_B_barcodes.tsv.gz		
GSM6908656_Day7_B_features.tsv.gz		
GSM6908656_Day7_B_matrix.mtx.gz		
GSM6908657_Day10_A_barcodes.tsv.gz		
GSM6908657_Day10_A_features.tsv.gz		
GSM6908657_Day10_A_matrix.mtx.gz		
GSM6908658_Day10_B_barcodes.tsv.gz		
GSM6908658_Day10_B_features.tsv.gz		
GSM6908658_Day10_B_matrix.mtx.g		
```


# Data Load


```{r}
data_dir <- '/Users/s2992547/data_local/datasets/GSE221883_DigitsDish_ScRNAseq/data_for_seurat/'
seurat_objects_dir <- '/Users/s2992547/data_local/datasets/GSE221883_DigitsDish_ScRNAseq/seurat_objects/'
```


```{r}
samples <- list.files(data_dir)

sample_dirs <- file.path(data_dir, samples)
names(sample_dirs) <- samples

data <- Read10X(data.dir = sample_dirs)
so   <- CreateSeuratObject(counts = data, project = "Fuiten2023")

so[["percent.mt"]] <- PercentageFeatureSet(so, pattern = "^mt-")


```

```{r}
VlnPlot(so, features = c("nFeature_RNA"))
VlnPlot(so, features = c("nCount_RNA")) + scale_y_log10()
VlnPlot(so, features = c("percent.mt"))
```

Add basic sample information.
```{r}
anno_table <- as_tibble(str_split_fixed(rownames(so@meta.data), "_", n = 4 ))
colnames(anno_table) <- c("Accession", "Day","Rep","Cell")
so[["Sample"]] <- paste(anno_table$Day, anno_table$Rep, anno_table$Accession, sep="_")
so[["Accession"]] <- anno_table$Accession
so[["Day"]] <- anno_table$Day
so[["Rep"]] <- anno_table$Rep
so[["Cell"]] <- anno_table$Cell
```


Do routine processing (absolutely not optimised for this study, just need something reasonable.)
```{r}
so <- subset(so, subset = nFeature_RNA > 2000 & percent.mt < 25)
so <- NormalizeData(so)
so <- FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000)
so <- ScaleData(so) # No cc regression.
so <- RunPCA(so, features = VariableFeatures(object = so))
so <- RunUMAP(so, dims = 1:20)
so <- FindNeighbors(so, dims = 1:20)
so <- FindClusters(so)
```


```{r}
ElbowPlot(so,ndims = 30)
```


UMAP views
```{r}
DimPlot(so)
DimPlot(so, group.by='Sample')
DimPlot(so, group.by='Day')
DimPlot(so, split.by='Sample', ncol=3)
```

```{r eval=FALSE}
saveRDS(so, file.path(seurat_objects_dir,"Fuiten2023_DigitsInDish_00_load.RDS"))
```



