---
title: "Differential celltype composition between groups"
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---



Is there a difference in the celltype composition between individuals with  Ulcerative colitis  or Crohn's disease, and Healthy controls?


# Load libraries and data object

```{r}
library(Seurat)
library(speckle)
library(tidyverse)
```

```{r}
data_dir              <- file.path("~/data_local/datasets/GSE234713_IBDcosmx_GarridoTrigo2023/processed_data") 
seurat_file_01_loaded <- file.path(data_dir, "GSE234713_CosMx_IBD_seurat_01_loaded.RDS")
```


```{r}
so <- readRDS(seurat_file_01_loaded)

# MIGRATE TO DATA PREP >>>>
so$individual_code <- factor(substr(so$orig.ident,12,16))
so$tissue_sample   <- factor(substr(so$orig.ident,12,16))
so$fov_name        <- paste0(so$individual_code,"_", str_pad(so$fov, 3, 'left',pad='0'))
so$celltype_subset <- factor(so$celltype_subset)
# <<<<
```

# Looking at the data

<!-- Move this to own page? --->

There are three indivduals per contidion (one tissue sample from each individual). With multiple fovs on each physical tissue sample.
```{r}
select(as_tibble(so@meta.data), condition, individual_code, fov_name) %>%
  unique() %>% 
  group_by(condition, individual_code) %>% 
  summarise(n_fovs= n(), item = str_c(fov_name, collapse = ", "))
```








# Full worked example 

## Count how many cells of each type in your data

```{r}
celltype_summary_table <- so@meta.data %>% 
  group_by(condition, individual_code, fov_name, celltype_subset) %>%
  summarise(cells=n(), .groups = 'drop')

```

## check for very low count groups


Here T cells are rare, but there are still a decent distribution of them with 10-100+ cells in a FOV.

If many of your celltypes, consider merging similar cell types (e.g. T cells rather than T cell subtypes)


```{r}
ggplot(celltype_summary_table, aes(x=cells, col=celltype_subset)) +
  geom_density() +
  geom_rug(alpha=0.2) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Cells per FOV by celltype")
```


-----


If you have alot of cell types, sometimes there can be very rare types that would be hard to detect differences in. 

This can expecially happen if you're using celltype assignment with a detailed reference. You might get a handful of irrelevant cell types called (e.g. 4 hepatocytes on a non-liver sample). No reasonable stats could be generated there, and leaving them in would mean a more extreme FDR adjustment. 



```{r}
celltype_summary_table.SingleR <- so@meta.data %>% 
  group_by(condition, individual_code, fov_name, celltype_SingleR2) %>%
  summarise(cells=n(), .groups = 'drop')



ggplot(celltype_summary_table.SingleR, aes(x=cells, col=celltype_SingleR2)) +
  geom_density() +
  geom_rug(alpha=0.2) +
  scale_x_log10() +
  theme_bw() +
  ggtitle("Cells per FOV by celltype")
```


## Look at your samples. 


```{r}
ggplot(celltype_summary_table, aes(x=fov_name, y=cells, fill=celltype_subset)) +
  geom_bar(position="fill", stat="identity") + 
  theme_bw() +
  coord_flip() + 
  theme(legend.position = "bottom") +
  facet_wrap(~condition, ncol=3, scales = 'free_y') +
  scale_y_continuous(expand = c(0,0)) 
```


## Calculate stats.


```{r}
results.anova <- propeller(clusters= so$celltype_subset, 
          sample  = so$individual_code, 
          group   = so$condition)
results.anova

# If a column is preferred over rownames
results.anova.table <- rownames_to_column( results.anova, var="celltype_subset")
```


```{r}
so.UCvsHC <- so[,so$condition %in% c("Healthy controls", "Ulcerative colitis")]


results.pair <- propeller( clusters= so.UCvsHC$celltype_subset, 
                           sample  = so.UCvsHC$individual_code, 
                           group   = so.UCvsHC$condition)
```

# Code snippet


```{r eval=FALSE}
library(speckle)
# seurat object so
results_table <- propeller(clusters = so$cluster, 
                           sample   = so$sample, 
                           group    = so$condition)


```

# Results


```{r}
results.anova
```

* rownames : 
* BaselineProp 
* PropMean.Chrones.s.disease 
* PropMean.Healthy.controls 
* PropMean.Ulcerative.colitis
* Fstatistic   
* P.Value
* FDR

# More information

* Propeller paper: https://academic.oup.com/bioinformatics/article/38/20/4720/6675456





