---
title: "Neighbourhood changes"
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6096346/


https://github.com/lfranzen/scwat-st.git


```{r angrylibrarian, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(echo = TRUE)
```

<!-- keywords

insituspatial
seurat
test
neighbours
-->


# Overview

## Load libraries and data

```{r}
library(Seurat)
library(tidyverse)
library(DT)
library(patchwork)
```

```{r}
data_dir              <- file.path("~/projects/spatialsnippets/datasets/GSE234713_IBDcosmx_GarridoTrigo2023/processed_data") 
seurat_file_01_loaded <- file.path(data_dir, "GSE234713_CosMx_IBD_seurat_01_loaded.RDS")
seurat_file_02_niche <- file.path(data_dir, "GSE234713_CosMx_IBD_seurat_02_niche.RDS")

seurat_file_02_niche.mini <- file.path(data_dir, "GSE234713_CosMx_IBD_seurat_02_niche_mini.RDS")

```





# Worked Example (1)


## Load Data



```{r}
so <- readRDS(seurat_file_01_loaded)
```


## Get cell neighbours


When you compute neighbourhood niches, Seurat adds a 'niche' column to the cell metadata, 
but in order to calculate that it also builds the 'niche' Assay within the object.

This niche assay is a matrix of cell X celltype - counting how many cells 'neighbouring' each cell have which cell type. 


So even if you don't want the defined niches, an easy way to find cell neighbours is to calculate the niches. If you haven't already. 


NB:   neighbors.k = 20,

```{r eval=FALSE}
#NB Current version of Seurat doesn't build niche assay accross multiple slides.
# This code will do so. Hope its fixed in future version of Seurat. 
source("code/multislide_niche.R")
so <- BuildNicheAssay.using_all_fovs(so, group.by = "celltype_subset",niches.k = 6)
DefaultAssay(so) <- "RNA" # change default assay back to RNA to avoid confusion.


saveRDS(so , seurat_file_02_niche)
```
```{r}
# reload
so <- readRDS(seurat_file_02_niche)
```


This can take some time to run. 

Where the original *RNA* assay has ~1000 gene features, the *niche* assay has 5 features - one for each celltype. And that matrix contains for every cell the number of neighbouring cells it has of each cell type. 

There are 20 neighbours for each cell, because the BuildNicheAssay parameter *neighbours.k* was the default of 20.

```{r}
dim(GetAssay(so, assay="RNA"))
dim(GetAssay(so, assay="niche"))
GetAssayData(so, assay="niche", layer='counts')[,1:4]
```

## Filter

We might choose to exclude 'poor quality' fovs that have few cells. Plotted below, the remaining cells in those FOVs are spaced further apart, so the neighbourhood is less useful. The 200 here is arbitrary.

(Using a method methods that can use maximal distances would avoid this issue)


```{r}
min_cells_per_fov <- 200

cellcounts <- stack(table(so$fov_name))
colnames(cellcounts) <- c('n_cells','fov_name')

ggplot(cellcounts, aes(x=n_cells, ) )+
  geom_density() +
  geom_rug() + 
  geom_vline(xintercept = min_cells_per_fov, lty=2, col='red')+
  theme_bw() +
  ggtitle("Cells per fov")

```


Make a table of the fovs that have low counts
```{r}
cellcount_fov_info <- so@meta.data %>% 
  group_by(fov_name, orig.ident) %>%
  summarise(n_cells = n()) %>%
  arrange(n_cells)

DT::datatable(cellcount_fov_info)       
```

```{r fig.width=12, fig.height=10}
# A typical sample with reasonable cell counts;
# 1855 cells
so.this <- subset(so, fov_name == 'CD_c_006')
p1 <- ImageDimPlot(so.this, 
               fov = 'GSM7473690_CD_c', # seurat's fov is slide,
               group.by = 'celltype_subset', axes = TRUE)

# 1488 cells
so.this <- subset(so, fov_name == 'HC_a_019')
p2 <- ImageDimPlot(so.this, 
               fov = 'GSM7473682_HC_a', # seurat's fov is slide,
               group.by = 'celltype_subset', axes = TRUE)
p1+p2
  


# Versus some of these low counts
	
# 95 cells
so.this <- subset(so, fov_name == 'HC_c_012')
p1 <- ImageDimPlot(so.this, 
               fov = 'GSM7473684_HC_c', # seurat's fov is slide,
               group.by = 'celltype_subset', axes = TRUE)

# 194 cells
so.this <- subset(so, fov_name == 'CD_a_016')
p2 <-ImageDimPlot(so.this, 
               fov = 'GSM7473688_CD_a', # seurat's fov is slide,
               group.by = 'celltype_subset', axes = TRUE)

p1 + p2
```




Apply filter

```{r}
#low_count_fov_info <- filter(cellcount_fov_info, n_cells < min_cells_per_fov) 
keep_fov <- filter(cellcount_fov_info, n_cells >= min_cells_per_fov) %>% pull(fov_name)
so <- subset(so, cells = colnames(so)[so$fov_name %in% keep_fov])
```







## Plotting



### Just one relationship

Are there more t-cells near epithelia in different conditions?


```{r}
plottable <- FetchData(so, vars=c('group', 'individual_code', 'fov_name', 'celltype_subset', "tcells")) %>% filter(celltype_subset == "stroma")
head(plottable)
```

The distribution shown for individual cells.
```{r}
ggplot(plottable, aes(x=individual_code, y=niche_tcells, col=group)) +
  geom_violin() +
  facet_wrap(~group, scales = "free_x") +
  theme_bw()
```



And showing averages per fov, grouped by individual sample.
```{r}
plottable2 <- plottable %>% 
  group_by(group, individual_code, fov_name, celltype_subset) %>%
  summarise(mean_t_cell_neighbour = mean(niche_tcells))
```
```{r}
ggplot(plottable2, aes(x=individual_code, y=mean_t_cell_neighbour, col=group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  facet_wrap(~group, scales = "free_x") +
  theme_bw()
```

### All vs All Celltypes

Or, more broadly

Note the warning, the FetchData function looks in the cell metadata for these columns, then the default assay (RNA), and then in other assays, finally finding tcells as a column there.
```{r}
celltypes <- levels(so$celltype_subset)
plottable3 <- FetchData(so, vars=c('group', 'individual_code', 'fov_name', 'celltype_subset', celltypes)) %>%
  pivot_longer(cols = starts_with("niche_"), names_to = "neighbour_celltype", values_to = "n_neighbours") %>%
  mutate(neighbour_celltype = gsub("niche_","", neighbour_celltype))
  

plottable4<- plottable3 %>% 
  group_by(group, individual_code, fov_name, celltype_subset, neighbour_celltype) %>%
  summarise(mean_neighbours = mean(n_neighbours)) %>%
  mutate(
    celltype_subset    = paste0(celltype_subset, " cells"),
    neighbour_celltype = paste0(neighbour_celltype, " neighbours"),
  )

# or by bio replicate
plottable5<- plottable3 %>% 
  group_by(group, individual_code, celltype_subset, neighbour_celltype) %>%
  summarise(mean_neighbours = mean(n_neighbours)) %>%
  mutate(
    celltype_subset    = paste0(celltype_subset, " cells"),
    neighbour_celltype = paste0(neighbour_celltype, " neighbours"),
  )


```


```{r fig.width=12, fig.height=10}
ggplot(plottable4, aes(x=group, y=mean_neighbours)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(mapping=aes(col=individual_code), shape=4) +
  facet_grid(neighbour_celltype ~ celltype_subset) +
  theme_bw()
```


```{r fig.width=12, fig.height=10}
ggplot(plottable5, aes(x=group, y=mean_neighbours)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  facet_grid(neighbour_celltype ~ celltype_subset) +
  theme_bw()
```

## Test

<!-- I should email and ask if this is a suitable application of propellar -->

>❓**_NOTE:_**  Can you suggest a better approach? Please get in touch.


To formally test this - Do we see more T cells near epithelial cells? 
Or, conversly, more eipithelial cells near T cells? (Which might be a different question if the T cells are doing something else.)

If so, we expect the proportions of cells in the neighbourhood to change.

Propeller can be use to test changes in cell type proportions between groups.


**NB: Work in progress - Needs statistical review.**


```{r}
library(limma)
library(speckle)

# Unit
so$fov_celltype <- paste0(so$fov_name, "_", so$celltype_subset)

# Get counts of neighbourhood for each cluster+fov group
neighbour_counts <- AggregateExpression(so, 
                                        assays = "niche", data="counts", group.by = "fov_celltype")
neighbour_counts_matrix <- neighbour_counts$niche
colnames(neighbour_counts_matrix) <- gsub("-","_", colnames(neighbour_counts_matrix))
#"UC-c-015-epi vs HC_a_001_epi"


# Total number of neibhours of celltypes of celltype X per fov
# (always 20 total)
total_neighbour <- colSums(neighbour_counts_matrix)
# Proportions of total
neighbour_proportions <- t(t(neighbour_counts_matrix) / total_neighbour)

# from the defulat of propeller.
neighbour_proportions.arcsine <- asin(sqrt(neighbour_proportions ))



# Make a table of relevant sample information, in same order as props, and check.
sample_info_table <- so@meta.data %>% 
  select(fov_celltype, fov_name, celltype_subset, tissue_sample, group) %>%
  unique()



row_order <- match(colnames(neighbour_proportions.arcsine),sample_info_table$fov_celltype)
sample_info_table <- sample_info_table[row_order,]
stopifnot(all(sample_info_table$fov_celltype == colnames(neighbour_proportions.arcsine))) # check it


```


```{r}

# Filter to just *t-cells*, in the sample table. 
# where neibhours could be anything.
celltype_to_test <- 'tcells'
sample_info_table.this <- sample_info_table[sample_info_table$celltype_subset == celltype_to_test, ]
#transformProps.this    <- props$TransformedProps[,sample_info_table.this$fov_celltype]
neighbour_proportions.arcsine.this <- neighbour_proportions.arcsine[,sample_info_table.this$fov_celltype]


# Extract relevant factors in same order as props
sample          <- sample_info_table.this$tissue_sample
group           <- sample_info_table.this$group
design <- model.matrix( ~ 0 + group)

dupcor <- duplicateCorrelation(neighbour_proportions.arcsine.this, design=design,  block=sample)

fit <- lmFit(neighbour_proportions.arcsine.this, design=design, block=sample, correlation=dupcor$consensus)

# Contrast called 'test', measuring of test condition vs Control condition.
contrasts <- makeContrasts(
  CDvsHC= groupCD - groupHC,
  UCvsHC= groupUC - groupHC,
  levels = coef(fit))
fit <- contrasts.fit(fit, contrasts)
fit <- eBayes(fit)

results_table <- topTable(fit, coef='CDvsHC')


```

```{r}
DT::datatable(results_table )
```




# Code Snippet


# Results


# More Information


# Scratch


```{r eval=FALSE}

so <- readRDS(seurat_file_01_loaded)

so <- subset(so, fov_name %in% c("HC_a_001", 'HC_b_001','HC_c_001',
                                 "CD_a_001", "CD_b_016","CD_c_001"))


#NB Current version of Seurat doesn't build niche assay accross multiple slides.
# This code will do so. Hope its fixed in future version of Seurat. 
source("code/multislide_niche.R")
so <- BuildNicheAssay.using_all_fovs(so, group.by = "celltype_subset",
                                     niches.k    = 6,
                                     neighbors.k = 5) #only 5 neighbour cells.
DefaultAssay(so) <- "RNA" # change default assay back to RNA to avoid confusion.


#saveRDS(so , seurat_file_02_niche.mini)
```
```{r}
# reload
so <- readRDS(seurat_file_02_niche)
```


Here we have neightoubing cell types per cell.

Will we be able to run a nighbouring cell type assy per cell or will we need to do ig 
```{r}
GetAssayData(so, assay="niche", layer='counts')[,1:4]
```


Semla (successor of STutility) is the framework. Doesn't seem to have an explicity funciton for neighberhood, so adapt the methods provided by the paper [Spatial mapping reveals human adipocyte subpopulations with distinct sensitivities to insulin](https://www.sciencedirect.com/science/article/pii/S1550413121003636) (@backdahlSpatialMappingReveals2021), in their repo: https://github.com/lfranzen/scwat-st.git


Their heterotypic score algorithm is here:
https://github.com/lfranzen/scwat-st/blob/master/scripts/visium_nbs_analysis.between-clusters.R 

```{r}
# Adapting here.
source('code/scwat-at_betweenClusters_heterotypic_score_functions.R')
```

```{r}
#' "EXPECTED" VALUES: RegionNeighbours() with permuted cluster IDs for each sample
n_perm <- 3 # 50
perm_adj_mat_list <- list()
c_include <- levels(so$celltype_subset)
#i=1
for(i in 1:n_perm){
  so <- RandomiseClusteringIDs(se.object = so, 
                               column.cluster.id = "celltype_subset", 
                               column.sample.id  = "tissue_sample", 
                               random.seed = i)
  so <- SetIdent(so, value = "clusters_perm")
  #for(column_rm in grep(pattern = "nbs_", colnames(so[[]]), value = T)){
  #  so[[column_rm]] <- NULL
  #}
  for(c in c_include){
    so <- RegionNeighbours(so, id = c, keep.within.id = T, verbose = TRUE)
  }
  perm_nbs_df <- so[[]][, c("clusters_perm", grep(pattern = "nbs_", colnames(so[[]]), value = T))]
  perm_adj_mat <- CreateAdjMatrix(nbs.df = perm_nbs_df, cluster.include = c_include, column.clusters.id = "clusters_perm")
  perm_adj_mat_list[[i]] <- perm_adj_mat
}


```





```{r}
library(igraph)

neighbors.k <- 5


## Get neigherest neibhtours Graph

fov <- names(so@images)[1]

coords <- SeuratObject::GetTissueCoordinates(so, image=fov, which = "centroids")
rownames(coords) <- coords$cell
coords      <- as.matrix(coords[ , c("x", "y")])
neighbors    <- Seurat::FindNeighbors(coords, k.param = neighbors.k, return.neighbor=TRUE)


knn <- BiocNeighbors::findKNN(coords, k=5)
class(knn[["index"]])
 
 
 
## Get Cell type adjacency 

## Use label shuffling to generate the randomised cell tyep adjacencies





cell_adj_matrix <- neighbors$nn
cluster_lookup <- so$celltype_subset
cluster_list <- levels(so$celltype_subset)



# Designed to be called 50 times with shuffled labels, it only recounts. 
celltype_adjmatrix_from_cell_adjmatrix_and_labels <- c(cell_adj_matrix, cluster_list, clusters_lookup) {
  #clust_name <- 'epi'
  
  sum_rows_by_type <- function(clust_name) {
    rowSums(cell_adj_matrix[,cluster_lookup[colnames(cell_adj_matrix)] %in% clust_name])
  }

  # have csll X cell adjacentcy matrix
  # For row cells, summarise columns into sums by cell type
  celltype_neighbours.1 <- sapply(sum_rows_by_type, X=cluster_list )
  
  # Have cell X celltype matrix
  # For column celltypes, summarise rows into sums by cell type.
  sum_cols_by_type <- function(clust_name) {
    colSums(celltype_neighbours.1[cluster_lookup[rownames(celltype_neighbours.1)] %in% clust_name,])
  }

  celltype_adj_matrix <- sapply(sum_cols_by_type, X=cluster_list )
  return(celltype_adj_matrix)
}





```



Get adjacency matrix? Of celltype => celltype.
From cell => adjcent celtype.
```{r}
library(igraph)
 neighbors.k <- 5


 
 
  
 
 
 
 
 
 
 
 get_nn_spatial_graph <- function(object) {
 
  for (fov in names(object@images)) {
    coords <- GetTissueCoordinates(object[[fov]], which = "centroids")
    rownames(coords) <- coords$cell
    
    coords <- as.matrix(coords[ , c("x", "y")])
    neighbors    <- FindNeighbors(coords, k.param = neighbors.k)
    neighbors$nn <- neighbors$nn[Cells(object[[fov]]), Cells(object[[fov]])]

    # Need a list of all those cells, used later.
    # put all cells into one list (being sure order matches network, might not match object order.)
    cells <- colnames(neighbors$nn)
    
    
  }
}

   
   
class(neighbors$nn)
#make_adjmatrix_graph(neighbors$nn, directed = FALSE)

 
 
 
 
#Just change labels and use the same matrix
# no can't use niche because that's the right  celltypess
#celltype_adjmatrix_from_niche_assay <- function(so, cluster_col) {
#  GetAssayData(so, "niche")
#}


 
 
 

#niche_assay <- GetAssayData(so, assay="niche", layer='counts')
 
 fov <- "GSM7473684_HC_c"
so.this <- subset(so, fov==fov)

#for (fov in names(so@images)) {
  coords <- GetTissueCoordinates(so.this[[fov]], which = "centroids")
  rownames(coords) <- coords$cell
  coords <- as.matrix(coords[ , c("x", "y")])
#}




neighbors    <- FindNeighbors(coords, k.param = neighbors.k)
#neighbors$nn <- neighbors$nn[Cells(object[[fov]]), Cells(object[[fov]])]


# rows and columns are cells.


# the nn graph is based on cell cell adjacency. 
# The snn graph is not, that's transcriptomic similarity?
adj_matrix <- neighbors$nn 






```













# References


