---
title: "Getting Compute"
author: "Sarah Williams"
output:
  html_notebook: 
    toc: yes
    code_folding: hide
  html_document:
    toc: yes
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---



## Getting a VM with R and Rstudio




## Installing R packages


These can be fast effienct C libraries that R uses in the background to handle number crunching, file format processing, other programming languages or programs. 



### Suggested system libraries


To save time, consider installing the following before trying to install R packages

```{sh eval=FALSE}


```


### Installing R packages


Suggested package installs for a new setup - You'll likely want more later!

These's installations will takes some time to run (by default they'll compile from source).

 * Might stop respongind - just reconnect if needed, the command can be rerun and will pcik up from where it left.
 * Might throw some errors - see next section;

Install some R packages from CRAN;

```{r}

install.packages ( c(
  "BiocManager",
  'tidyverse',
  'DT',
  'patchwork',
  'ggrepel'
  ) )

```


Install some R packages from bioconductor

```{r}
BiocManager::install(c(
  'alabaster.sfe', 
  'SpatialExperiment',
  'scuttle',
  'scater',
  'limma',
  'edgeR',
  'BiocParallel',
  'Voyager'
))
```



Install a few things from github.

```{r}
devtools::install_github("immunogenomics/presto") # Dramatic speed up of certain operations in Seurat.
```






### Its not working! Handling R package install errors



If you get an error in install


Look for 'depencancy 'x' was not installed. Try to install package 'x' on its own - you'll get the error much quicker and can fix it. 


Issues

* Missing system library
* System library configuration issue (e.g. rJava)
* Its a bioconductor package (install.packages can't see it)
* Its not in CRAN either - install from a github repo
* The package you want is on a newer version of bioconductor. 


Check out the 

https://bookdown.org/guokai8/mastering-r-through-errors/docs/installation-environment.html#debugging-installation-issues









```
...
Error: package ‘SpatialFeatureExperiment’ 1.4.0 was found, but >= 1.9.3 is required by ‘alabaster.sfe’
Execution halted
....
```
I am installing alabaster.sfe form github, which expects the latest version of SpatialFeatureExperiment




### Upgrading R

Bioconductor has a 6-monthly release cycle, and strict versioning requirements. Each bioconductor version has a minimum R requirement - and for the very latests this could be less than what is on bioshell!

If you want to work with packages (or versions) only available in 


|Release |	Date 	        |  	R  |
|------|------------------|------|
| 3.23 | April 29, 2026   |  4.6 |
| 3.22 | October 30, 2025 |  4.5 |
| 3.21 | April 16, 2025   |  4.5 |
| 3.2  | October 30, 2024 |  4.4 |
| 3.19 | May 1, 2024      |  4.4 |
| 3.18 | October 25, 2023 |  4.3 |


```{sh eval=FALSE}
 shelley build r/4.6.1
```







## Getting the data there.


### Example datasets



```{sh eval=FALSE}
mkdir data
cd data
#wget --content-disposition "https://filesender.aarnet.edu.au/download.php?token=378da83e-7c11-40d8-a90c-eddb5da9c109&files_ids=29832957"
# wget <nectar object store link>
tar -xzf GSE234713_CosMx_IBD_sfe_03_banksyniche.tar.gz
```



