---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---



These examples are a sampler of tests that can be performed on an in-situ single cell spatial datasets. Initial  examples use cosMX, though approaches should be valid on other platforms like Xenium. 

Further details on the papers, tools, statistics and methods are linked from each page.

Each example presents just one way of doing things (and currently only with Seurat) - there are many, many other approaches! These are intended as a starting point.

These examples assume that the data has been loaded, preprocessed (e.g. QC-ed, UMAP) and annotated (e.g. sample names, cell types, cell niches). For information on how each dataset was prepared, see the data tab. 


```{r echo=FALSE, warning=FALSE, message=FALSE}
library(reactable)
library(htmltools)
library(dplyr)

howto_info_file <- '/home/s.williams/projects/spatialsnippets/spatialsnippets/data/howto_info.tsv'
howto_info <- read.delim(howto_info_file, sep="\t")
howto_info$image <- paste0("assets/",howto_info$image)
```


```{r echo=FALSE}
#Defaine column widths;
#https://glin.github.io/reactable/articles/examples.html#column-widths

# reactable leaves col order as is.
data <- select(howto_info, image, howto, Description,  Technology, Toolkit, url, everything() )

reactable(
  data,
  
  searchable = TRUE, # filtering is also possible, but at this stage, a simple search box is enough.
  
  columns = list(
    
    image = colDef(
      name = "",
      minWidth = 250,
      cell = function(value, index) {
        
        image <- img(src = value, 
                     #style = "object-fit: contain",  
                     alt = value
                     ) #
        link_with_image <- tags$a(
            href = data$url[index],
            image
        )
        
        
        tagList(
          tags$a(style = "font-weight: 600; color:black;", href = data$url[index], target = "_blank",  data$howto[index]),
          div(style = "display: inline-block;", link_with_image)

        )
        
      }),
    
    howto       = colDef(show = FALSE),
    
    Description = colDef(minWidth = 200),

    # By default hidden columns aren't searched, even though the content is copied
    # into a different column. 
    # Searchable=TREU  enables search of specific tech, e.g. 'cosmx' to work
    SpecificTech     = colDef(show = FALSE, 
                              searchable = TRUE), 
    

    #Combine broad and specific tech    
    Technology = colDef(
      cell = function(value, index) {
          specificTech <- data$SpecificTech[index]
          div(
            div(value),
            div(style = "font-style: italic", specificTech)
          )
    }),
    
    Toolkit = colDef(minWidth = 50),
    
    
    url = colDef(
      name = "",
      width = 50,
        cell = function(value) {
          tags$a(href = value, target = "_blank",  "▶️")
      }
      
    )
  )
    
)



```








