---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
bibliography: /home/s.williams/zotero_export/MyLibrary.bib
---

Below is a sampler of tests that can be performed on an in-situ single cell spatial datasets. Each example presents just one way of doing things - there are many, many other approaches! These are simply a starting point.



```{r echo=FALSE, warning=FALSE, message=FALSE}
library(reactable)
library(htmltools)
library(dplyr)

howto_info_file <- '/home/s.williams/projects/spatialsnippets/spatialsnippets/data/howto_info.tsv'
howto_info <- read.delim(howto_info_file, sep="\t")
howto_info$image <- paste0("assets/",howto_info$image)
```


```{r echo=FALSE}
#Defaine column widths;
#https://glin.github.io/reactable/articles/examples.html#column-widths

# reactable leaves col order as is.
data <- select(howto_info, image, howto, Description,  Technology, Toolkit, url, everything() )

reactable(
  data,
  
  searchable = TRUE, # filtering is also possible, but at this stage, a simple search box is enough.
  
  columns = list(
    
    image = colDef(
      name = "",
      minWidth = 250,
      cell = function(value, index) {
        
        image <- img(src = value, 
                     #style = "object-fit: contain",  
                     alt = value
                     ) #
        link_with_image <- tags$a(
            href = data$url[index],
            image
        )
        
        
        tagList(
          tags$a(style = "font-weight: 600; color:black;", href = data$url[index], target = "_blank",  data$howto[index]),
          div(style = "display: inline-block;", link_with_image)

        )
        
      }),
    
    howto       = colDef(show = FALSE),
    
    Description = colDef(minWidth = 200),

    # By default hidden columns aren't searched, even though the content is copied
    # into a different column. 
    # Searchable=TREU  enables search of specific tech, e.g. 'cosmx' to work
    SpecificTech     = colDef(show = FALSE, 
                              searchable = TRUE), 
    

    #Combine broad and specific tech    
    Technology = colDef(
      cell = function(value, index) {
          specificTech <- data$SpecificTech[index]
          div(
            div(value),
            div(style = "font-style: italic", specificTech)
          )
    }),
    
    Toolkit = colDef(minWidth = 50),
    
    
    url = colDef(
      name = "",
      width = 50,
        cell = function(value) {
          tags$a(href = value, target = "_blank",  "▶️")
      }
      
    )
  )
    
)



```



## Further information


More information on the source datasets may be found on the data tab.

Initial examples use cosMX data, though approaches should be valid on other platforms like Xenium.

Initial examples use the seurat toolkit, but single feature experiment (bioconductor) examples are in active development. 
